Avoiding deforestation: The implications for global markets

### 2024-02-28 paper.Rmd
Update paper.Rmd

---
title: "Implications of afforestation policies for global markets"
date: "`r format(Sys.time(), '%B %d, %Y')`"
author: "Eric C. Davis, Maros Ivanic, and Brent Sohngen"
output: 
  bookdown::pdf_document2:
    keep_tex: TRUE
toc: FALSE
bibliography: references.bib
header-includes:
    - \usepackage{setspace}\doublespacing
abstract: "The projected growth in population and incomes as implied by various shared socioeconomic pathways is expected to create pressure to convert forestland into farmland. At the same time, the increasingly negative climate impacts are expected to generate increasing pressure to enhance the terrestrial carbon sink. These goals seem incompatible as reversing the deforestation trend by afforesting cropland might result in negative market impacts such as higher food prices, especially in low agricultural productivity/high population growth scenarios. In order to better grasp the interrelationship and the magnitude of harm that might be caused by prioritizing the goal of mitigating deforestation, this analysis connects two respected global economic models: the Global Trade Analysis Project (GTAP) model and the Global Timber Model (GTM). Results from the model suggest that it is possible through fiscal policy to preserve 144.2 million hectares of forestland that otherwise would be converted to agricultural land by 2033. There does appear to be a rather significant economic price for doing so, as the avoided deforestation would result in less agricultural output and higher market prices. This is estimated to impact the well-being of global consumers by $92.1 billion, which translates to an global average cost of $10.6 per person in 2033."
---

@@

This analysis shows that, at an estimated tax rate of 70.3 percent of the value of each acre of land that is converted to agricultural land, global net deforestation would drop to roughly zero. In the United States, as USDA-NASS (Land Values 2022 Summary (August 2022)) valued the average acre of farmland in 2022 at $3,800, the tax would be $2,671.4/acre. In Brazil, CEIC valued the average price of Brazilian land at roughly 2,500 Brazilian Reals, which would put the tax at roughly $203 per acre. Governments would very likely be wary of farmer discontent were they to apply, but they would also have to take into account the larger impacts on their economies. Restricting agricultural land, as shown above, would for the most part restrict output and increase prices. It is uncertain whether the gains from the increase in prices would be sufficient to offset the loss of potential revenue that would have come from expanded agricultural operations. As governments might be reluctant to impose such taxes on their own citizens, it is possible that policymakers who wished to see deforestation stopped could impose import tariffs equal to the tax rate to achieve the same result.

>>>>

---
title: "Avoiding deforestation: The implications for global markets"
date: "`r format(Sys.time(), '%B %d, %Y')`"
author: "Eric C. Davis, Maros Ivanic, and Brent Sohngen"
output: 
  bookdown::pdf_document2:
    keep_tex: TRUE
toc: FALSE
bibliography: references.bib
header-includes:
    - \usepackage{setspace}\doublespacing
abstract: "The projected growth in population and incomes as implied by various shared socioeconomic pathways is expected to create pressure to convert forestland into farmland. At the same time, the increasingly negative climate impacts are expected to generate increasing pressure to enhance the terrestrial carbon sink. These goals seem incompatible as reversing the deforestation trend by afforesting cropland might result in negative market impacts such as higher food prices, especially in low agricultural productivity/high population growth scenarios. In order to better grasp the interrelationship and the magnitude of harm that might be caused by prioritizing the goal of mitigating deforestation, this analysis connects two respected global economic models: the Global Trade Analysis Project (GTAP) model and the Global Timber Model (GTM). Results from the model suggest that it is possible through fiscal policy to preserve 144.2 million hectares of forestland that otherwise would be converted to agricultural land by 2033. As to the economic price for doing so, the avoided deforestation would in most regions of the world result in less agricultural output and higher market prices. This is estimated to impact the well-being of global consumers by $92.1 billion, which translates to an global average cost of $10.6 per person in 2033."
---

@@

This analysis shows that, at an estimated tax rate of 70.3 percent of the value of each hectare of land that is converted to agricultural land, global net deforestation would drop to roughly zero by 2033, preserving 144.2 million hectares of forestland that otherwise would have been converted to agricultural land. In the United States, as USDA-NASS (Land Values 2022 Summary (August 2022)) valued the average hectare of farmland in 2022 at $1,537.8, the tax would equate to $1,081.1/hectare. In Brazil, CEIC valued the average price of Brazilian land at roughly 2,500 Brazilian Reals, which would put the tax at roughly $82.2 per hectare. The economic price of this would likely be transfered to consumers through less agricultural output and higher market prices. Overall, such an action is estimated to reduce consumer well-being by $92.1 billion or $10.6 per every person in 2033. Governments might be wary of farmer discontent were they to take such fiscal actions and thus might be reluctant to impose such taxes on them. Thus, policymakers who wish to see deforestation stopped might have to resort to import tariffs on products produced on land converted from forestland for agricultural use.
